Bayesian Deep Gaussian Processes for Correlated Functional Data: \\ A Case Study in Cosmological Power Spectra

### 2025-04-18 paper/main.tex
include some initial CAMB results; add basic cosmology details and references

\title{Bayesian Deep Gaussian Processes for Correlated Functional Data: \\
        A Case Study in Cosmological Power Spectra}
\author{Stephen A. Walsh\thanks{Corresponding author: Division of Natural Sciences, 
        Math and Technology, Elms College, {\tt walshst@elms.edu}} \and 
        Annie S. Booth\thanks{Department of Statistics, North Carolina State University} \and
        David Higdon\thanks{Department of Statistics, Virginia Tech} \and
        Marco A.R. Ferreira\footnotemark[3] \and
        Kelly Moran\thanks{Los Alamos National Laboratory} \and
        Katrin Heitmann\thanks{Argonne National Laboratory}}
\date{\today}

@@

We are particularly motivated by the Mira-Titan simulation suite \citep{moran2023mira}
which defines ``cosmologies'' according to eight cosmological parameters (\textcolor{red}{more on this
in Section TBD}).  For a particular cosmology, the simulation of the power matter spectra 
is rather complex.  Although we presume there exists some true matter spectra as a function
of wavenumber, we are not able to directly observe it.  Rather, the simulation outputs 
a convoluted patchwork of potential spectra, which vary in fidelity and which are trusted
only over particular wavenumber ranges.  Our first objective is to synthesize expert knowledge
and all of this simulation information into an effective estimate of the true underlying
matter spectrum.

@@

The Mira-Titan simulation suite is a dataset consisting of simulated power spectra for 
117 different cosmologies. Each of these cosmologies is specified by 8 cosmological 
parameter values, which we will denote as $\theta \in \mathbb{R}^8$. Different simulation 
output exists for different values of a ninth value, the redshift; in this work, we fix 
this value at 0. For each cosmology, we have a batch of 18 simulated spectra: an inexpensive 
spectrum estimated from linear perturbation theory ($y_p$), 16 spectra estimated from 
lower-resolutions simulations ($y_{\ell_r}, r \in \{1,\dots,16\}$), and one spectrum from 
a high resolution simulation ($y_h$).

@@

where $d=||x_i-x_j||^2$. From here, we estimate $\tau_\ell, \theta_\ell$ with maximum 
likelihood, and conditional on $\Sigma_\varepsilon$, we specify priors for $\tau_S, \theta_S$ 
following \cite{sauer2023active}: \textcolor{cyan}{Annie, please ensure the most up-to-date 
prior specs are shown above and here. Here is what I found from fit\$settings when using 
deepgp v1.1.3: $\theta_S \sim G(\alpha=1.2, \beta=1)$ and $\tau_S \propto 1/\tau_S^2$.}

@@

\textcolor{cyan}{Annie: want to add something here? For $\theta_w$, we also specify 
a Gamma prior of the form $\theta_w \sim G(\alpha=1.2, \beta=2)$. We discuss the 
elliptical slice sampling of $W$, but otherwise it is plugged in to the methods 
of the previous section with very few changes.}

@@

In addition to simulated data, we investigate model fits for different cosmological 
datasets from CAMB \citep{lewis2011CAMB}. \textcolor{magenta}{We have looked at two sets of model 
runs, ``LT" runs where there is an underlying ``true" value alongside 15 computer 
model runs, or the ``CAMB" runs, where we only have a single output. The performance 
from the CAMB runs was better (with a 6-dimensional design space), and also did not 
require obtaining a posterior mean from multiple model runs. When looking at the 
CAMB runs on the $\Delta$ space, we see it generally is withing $\pm 5\%$ error. 
Regarding these results, compared to the results for the 8-dimensional LT runs 
which had less impressive predictive performance, I'm open to suggestions on whether 
to include details on one or the other.}

\begin{figure}
    \centering
    \includegraphics[width=6in]{CAMB_fit_model1.jpg}
    \caption{Example of model fit for the first model from the CAMB data, with LOESS
             mean term subtracted. \textcolor{red}{Steve: Update this plot to be
             more clean, use different line types: dotted, dash, solid, etc.}}   
    \label{fig:fit_camb}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=6in]{plot_pca_CAMB.png}
    \caption{Proportion of $\Delta_{emu}$ vs. $\Delta_{holdout}$ for model 
             runs 11-42 (excluding 27-28).}   
    \label{fig:pca_camb}
\end{figure}

@@

\subsection{GP-PC Model}
\label{subsec:pca}

@@

In alignment with \cite{heitmann2009coyote} and \cite{higdon2010estcosmo}, we use 
GPs to model the principal components' weights; we shall denote this method GP-PC. 
We assume a powered exponential kernel for the Gaussian process and employ 
\texttt{GPfit} to perform a multi-start gradient based search for GP's hyperparameters 
\citep{macdonald2015gpfit}. That is, the $i$th vector of PC weights $w_i(\theta)$, 
$i \in \{1,\ldots,p_\eta\}$, will be modeled as a zero mean GP: 
$w_i(\theta) \sim GP(0, \sigma^2R)$, with

@@

We fix $\alpha=1.95$ and estimate each $\beta_k$ with $k \in \{1,...,p_\theta\}$. 
Using these trained GPs from the PCs, we predict the spectra for unobserved cosmologies.

>>>>

\title{Bayesian Deep Gaussian Processes for Correlated Functional Data: \\
        A Case Study in Cosmological Power Spectra}
\author{Stephen A. Walsh\thanks{Corresponding author: Division of Natural Sciences, 
        Math and Technology, Elms College, {\tt walshst@elms.edu}} \and 
        Annie S. Booth\thanks{Department of Statistics, North Carolina State University} \and
        David Higdon\thanks{Department of Statistics, Virginia Tech} \and
        Marco A.R. Ferreira\footnotemark[3] \and
        Jared Clark\footnotemark[3] \and
        Kelly Moran\thanks{Los Alamos National Laboratory} \and
        Katrin Heitmann\thanks{Argonne National Laboratory}}
\date{\today}

@@

We are particularly motivated by the Mira-Titan simulation suite \citep{moran2023mira}
which defines ``cosmologies'' according to eight cosmological parameters (more on this
in Section \ref{sec:data}).  For a particular cosmology, the simulation of the power matter spectra 
is rather complex.  Although we presume there exists some true matter spectra as a function
of wavenumber, we are not able to directly observe it.  Rather, the simulation outputs 
a convoluted patchwork of potential spectra, which vary in fidelity and which are trusted
only over particular wavenumber ranges.  Our first objective is to synthesize expert knowledge
and all of this simulation information into an effective estimate of the true underlying
matter spectrum.

@@

The Mira-Titan simulation suite is a dataset consisting of simulated power spectra for 
117 different cosmologies. Each of these cosmologies is specified by eight cosmological 
parameter values: matter density, baryon density, amplitude of density fluctuations, 
dimensionless Hubble parameter, spectral index of scalar perturbations,
dark energy equation of state parameters, and neutrino density. For more details 
on these parameters and their effects, see \cite{dodelson2020modern, aghanim2020planck, heitmann2016mira}. 
Parameter ranges for each of these parameters are specified in Table 2.1 of \cite{moran2023mira}.
For simplicity, we will denote these eight parameters as $\theta \in \mathbb{R}^8$. Different simulation 
output exists for different values of a ninth value, the redshift; in this work, we fix 
this value at 0. For each cosmology, we have a batch of 18 simulated spectra: an inexpensive 
spectrum estimated from linear perturbation theory ($y_p$), 16 spectra estimated from 
lower-resolutions simulations ($y_{\ell_r}, r \in \{1,\dots,16\}$), and one spectrum from 
a high resolution simulation ($y_h$).

@@

where $d=||x_i-x_j||^2$. From here, we estimate $\tau_\ell, \theta_\ell$ with maximum 
likelihood, and conditional on $\Sigma_\varepsilon$, we specify priors for $\tau_S, \theta_S$ 
following \cite{sauer2023active}: \textcolor{red}{Annie, please ensure the most up-to-date 
prior specs are shown above and here. Here is what I found from fit\$settings when using 
deepgp v1.1.3: $\theta_S \sim G(\alpha=1.2, \beta=1)$ and $\tau_S \propto 1/\tau_S^2$.}

@@

\textcolor{red}{Annie: want to add something here? For $\theta_w$, we also specify 
a Gamma prior of the form $\theta_w \sim G(\alpha=1.2, \beta=2)$. We discuss the 
elliptical slice sampling of $W$, but otherwise it is plugged in to the methods 
of the previous section with very few changes.}

@@

In addition to simulated data, we investigate model fits for different cosmological 
datasets from CAMB \citep{lewis2011CAMB}. These datasets contain power spectra based
on six-dimensional cosmological parameter settings. Similar to the Mira-Titan suite, the
CAMB dataset has multiple (sixty-four) batches of power spectra for each cosmology, with each batch 
containing multiple (fifteen) low-resolution spectra, and one high-resolution sprectum. 
Unlike the Mira-Titan suite, this dataset favorably contains an infinite-resolution 
spectrum, which can be treated as the true spectrum for each cosmology,
allowing us to accurately assess posterior mean estimates and credible intervals.

\begin{figure}
    \centering
    \includegraphics[width=6in]{CAMB_fit_model1.jpg}
    \caption{Example of model fit for the first model from the CAMB data, with LOESS
             mean term subtracted. \textcolor{red}{Currently, we can compare the 
             orange lines (posterior mean and 95\% credible interval) to the blue line,
             representing the infinite-resolution ``true" spectrum. Steve: Update this plot to be
             more clean, use different line types: dotted, dash, solid, etc.}}   
    \label{fig:fit_camb}
\end{figure}

We use 32 CAMB batches to train the DGP.HM model and reserve the other 32 as a test set.
To imitate the role of the perturbation theory output of Mira-Titan, we anchor estimates of the CAMB
power spectrum at the infinite-resolution for $\log_{10}(k) < -1.7$, and then rely 
strictly on the low- and high-resolution runs for $k$ above this. We take a weighted
average of the CAMB runs and assume heteroskedastic errors in a similar fashion, now with
no spatial dependence (compare low-resolution runs of Figure \ref{fig:fit_camb} 
with Figure \ref{fig:plot_data}).

@@

\subsection{Functional Prediction Model}
\label{subsec:pca}

@@

% This is old: switched to SEPIA
% We assume a powered exponential kernel for the Gaussian process and employ 
% \texttt{GPfit} to perform a multi-start gradient based search for GP's hyperparameters 
% \citep{macdonald2015gpfit}. 
In alignment with \cite{heitmann2009coyote} and \cite{higdon2010estcosmo}, we use 
GPs to model the principal components' weights. For direct comparison alongside
\cite{moran2023mira}, we employ a Python implementation of this approach called 
SEPIA \citep{gattiker2020sepia}. \textcolor{red}{Please note: currently show GPfit 
predictions for Mira-Titan Data, NOT SEPIA. The CAMB results are based on SEPIA.}
That is, the $i$th vector of PC weights $w_i(\theta)$, 
$i \in \{1,\ldots,p_\eta\}$, will be modeled as a zero mean GP: 
$w_i(\theta) \sim GP(0, \sigma^2R)$, with

@@

\textcolor{red}{We fix $\alpha=1.95$ and estimate each $\beta_k$ with $k \in \{1,...,p_\theta\}$. 
The above with $R_{ij}$ is based on GPfit; will update this to reflect SEPIA settings later.}
Using these trained GPs from the PCs, we predict the spectra for unobserved cosmologies.

\subsection{Predicting Spectra for CAMB Data}
\label{subsec:camb_pred}

From our 32 posterior means using DGP.HM, these are input into SEPIA in order to predict power spectra
for the 32 test cosmologies. With the infinite-resolution output, we have a sense of the ``best-case" 
predictions we can have (i.e., if we trained with the 32 true power spectra, as opposed to our 
estimated posterior means). In Figure \ref{fig:pca_preds_v_camb}, we see comparisons of the heldout 
infinite resolution alongside two predictions: one based on DGP.HM and the ``best-case" scenario using the
infinite-resolution spectra of the training set. There is a high level of agreement between the pairs
of predictions across cosmologies. Upon comparing MSEs for each of these prediction scenarios 
(Figure \ref{fig:mse_camb}, most are very similar across the two groups (with the two most difficult 
predictions being the exceptions), suggesting that overall DGP.HM is estimating the true spectra 
for each batch well.

\begin{figure}
    \centering
    \includegraphics[width=4in]{preds_diff1_4_8.jpg}
    \caption{Predictions for 16 of the 32 test cosmologies after centering and scaling. 
             The interpolated infinite resolution run is shown in black, the prediction 
             (trained on the infinite resolution) is in blue and the prediction from 
             the DGP.HM posterior mean is in red.}   
    \label{fig:pca_preds_v_camb}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=3in]{mse_dot.jpg}
    \caption{Comparison of MSE for CAMB predictions, when training with DGP.HM posterior 
             means (left), and training with the ``true" infinite-resolution spectra (right).}   
    \label{fig:mse_camb}
\end{figure}
